---
title: Data Preparation
author: Kartlyn Nguyen
format: typst
---

```{r}
# Load necessary libraries:
library(tidyverse)
```

```{r}
# Load Data:
citbi <- read_csv("data/citbi.csv")
citbi
```

```{r}
# Quickly explore data:
class(citbi) # "spec_tbl_df" "tbl_df"  "tbl"  "data.frame" 
typeof(citbi) # "list"

dim(citbi) # 30379 rows ~ 30379 patients and 26 columns
```


## Part I: Data Cleaning
**1. Address Missing Data**
- Missing values in this dataset are not only represented by NA, but also by specific coded values:
    a. 91: Pre-verbal / Non-verbal patients
    b. 92: Not Applicable (NA)

```{r}
# Investigate NA values:

# Take a look at columns with '91' values:
na_91 <- citbi |>
    select(Amnesia_verb, HA_verb)

# Then, take a closer look at Amnesia_verb column:
na_91|>
    group_by(Amnesia_verb) |>
    summarize(count = n()) #'91' has 10352 values and 'NA' has 1581 values

# Next, take a closer look at HA_verb column:
na_91 |>
    group_by(HA_verb) |>
    summarize(count = n()) #'91' has 9802 values and 'NA' has 450 values

# Finally, I convert all '91' values to NA values with ifelse:
citbi <- citbi |>
    mutate(
        Amnesia_verb = ifelse(Amnesia_verb == 91, NA, Amnesia_verb),
        HA_verb = ifelse(HA_verb, NA, HA_verb)
    )
# If the value is '91', convert it to NA; otherwise, keep the original value in the original column.
```

*Observation:* Now, both Amnesia_verb and HA_verb contain proper missing values (NA) instead of the placeholder code 91.

```{r}
# Same idea with columns that have '92' values: converted to NA.

na_92 <- citbi |>
    select(LocLen, SeizLen)

na_92 |>
    group_by(LocLen) |>
    summarize(count = n()) # 25586 '92' values & 1812 'NA' values

na_92 |>
    group_by(SeizLen) |>
    summarize(count = n()) # 29972 '92' values & 74 'NA' values

# Convert 92 values to NA values:
citbi <- citbi |>
    mutate(
        LocLen = ifelse(LocLen == 92, NA, LocLen),
        SeizLen = ifelse(SeizLen == 92, NA, SeizLen)
    )
```

```{r}
# Validation Check: to make sure there are no 91, 92 values in citbi
sum(citbi %in% c(91, 92), na.rm = TRUE) #Expected result: 0

# Result: 0. OKAY!
```

*Observation:* The data frame **citbi** no longer contains the coded values 91 or 92; these have been replaced with NA values to ensure proper handling of missing data.

**2. Change Variable Names**

```{r}
# Rename variables for better clarity and interpretability

# View current name:
citbi_column_names <- colnames(citbi)
citbi_column_names

# Better name, I think: 
# patient_number = PatNum, amnesia = Amnesia_verb , loss_of_consciousness_len = LocLen, seizure = Seiz, seizure_len = SeizLen, act_normal = ActNorm, headache = HA_verb, altered_mental_status = AMS, skull_fracture = SFxPalp, fontanelle_bulging = FontBulg, hematoma = Hema, clavicle_trauma = Clav, neurological_deficit = NeuroD, other_significant_injuries = OSI, citbi_outcome = PosIntFinal

citbi <- citbi |>
    rename( #dplyr function
        patient_number = PatNum, 
        amnesia = Amnesia_verb , loss_of_consciousness_len = LocLen, 
        seizure = Seiz, 
        seizure_len = SeizLen, 
        act_normal = ActNorm, 
        headache = HA_verb,
        vomit = Vomit,
        dizzy = Dizzy,
        altered_mental_status = AMS, 
        skull_fracture = SFxPalp, 
        fontanelle_bulging = FontBulg, 
        hematoma = Hema, 
        clavicle_trauma = Clav, 
        neurological_deficit = NeuroD, other_significant_injuries = OSI,
        ct_form = CTForm1,
        age_in_month = AgeInMonth,
        gender = Gender,
        ct_done = CTDone,
        death_tbi = DeathTBI,
        citbi_outcome = PosIntFinal
    )
# All other variable names are kept the same, including 4 columns: GCSEye, GCSVerbal, GCSMotor, and GCSTotal.
```

```{r}
# See new names:
colnames(citbi)
```

**3. Specify Variable Types**
```{r}
# Create a data frame that includes the data type of each column & na_count:

# to help me decide what to convert!
data.frame(
  type = sapply(citbi, class),
  na_count = sapply(citbi, function(x) sum(is.na(x)))
)

# 4 columns with na_count = 0: 'patient_number', 'GCSTotal', 'age_in_month', and 'ct_done' => Safe & clean: no conversion needed!
```

- All columns are now "numeric".

```{r}
# Convert yes/no columns to logical vectors with TRUE/FALSE answers
citbi <- citbi |>
    mutate(
        amnesia = as.logical(amnesia),
        seizure = as.logical(seizure),
        act_normal = as.logical(act_normal),
        headache = as.logical(headache),
        vomit = as.logical(vomit),
        dizzy = as.logical(dizzy),
        altered_mental_status = as.logical(altered_mental_status),
        skull_fracture = as.factor(skull_fracture), # treat 2 values (= unclear exam) as NA values
        fontanelle_bulging = as.logical(fontanelle_bulging),
        hematoma = as.logical(hematoma),
        clavicle_trauma = as.logical(clavicle_trauma),
        neurological_deficit = as.logical(neurological_deficit),
        other_significant_injuries = as.logical(other_significant_injuries),
        ct_form = as.logical(ct_form),
        ct_done = as.logical(ct_done),
        death_tbi = as.logical(death_tbi),
        citbi_outcome = as.logical(citbi_outcome)
    )
```

```{r}
# Convert NO yes/no columns (character vectors) to factors
# loss_of_consciousness_len, seizure_len, GCSEye, GCSVerbal, GCSMotor, and gender (6 columns total)

# Length of loss of consciousness
citbi <- citbi |>
    mutate(
    loss_of_consciousness_len = factor(loss_of_consciousness_len,
    levels = c(1,2,3,4),
    labels = c("<5 sec", "5 sec - <1 min", "1-5 min", ">5 min"))
    )

# Length of post-traumatic seizure
citbi <- citbi |>
    mutate(
        seizure_len = factor(seizure_len,
        levels = c(1,2,3,4),
        labels = c("< 1 min", "1 - <5 min", "5-15 min", ">15 min"))
    )

# GCS component: Eye
citbi <- citbi |>
    mutate(
        GCSEye = factor(GCSEye,
        levels = c(1,2,3,4),
        labels = c("None", "Pain", "Verbal", "Spontaneous"))
    )

# GCS component: Verbal
citbi <- citbi |>
    mutate(
        GCSVerbal = factor(GCSVerbal,
        levels = c(1,2,3,4,5),
        labels = c("None", "Incomprehensible sounds (moans)", "Inappropriate words (cries to pain)", "Confused (irritable/cries)", "Oriented (coos/babbles)"))
    )

# GCS component: Motor
citbi <- citbi |>
    mutate(
        GCSMotor = factor(GCSMotor,
        levels = c(1,2,3,4,5,6),
        labels = c("None", "Abnormal extension posturing", "Abnormal flexure posturing", "Withdraws to pain", "Localizes pain (withdraws to touch)", "Follow commands (spontaneous movement)"))
    )

# Gender
citbi <- citbi |>
    mutate(
        gender = factor(gender,
        levels = c(1,2),
        labels = c("Male", "Female"))
    )
```

{{< pagebreak >}}

## Part II: Exploratory Data Analysis (EDA)
- The goal of EDA is to better understand the data, identify patterns, and generate hypotheses.
- Additionally, EDA presents an opportunity to create data products that can be shared with others.

**4. Create a Missing Data Summary:**
- I used across() to compute missing-value counts and proportions for every variable, then reshaped the output into a tidy table using pivot_longer() and .value naming.

```{r}
missing_values <- citbi |>
    summarize(across(everything(),
    list(
        "n_missing" = ~sum(is.na(.)),
        "prop_missing" = ~mean(is.na(.))
    ),
    .names = "{.col}--{.fn}")   #Separate the column name with n_missing & prop_missing for clarity
    ) |>
    pivot_longer(               #Reshape from wide -> long
        cols = everything(),    #Select all columns
        names_to = c("variables", ".value"), #Split columns into "variables" (original variable name) and ".value" (creates new columns: n_missing, prop_missing)
        names_sep = "--"  #Match the separator above
    )
missing_values   # 28 rows & 3 columns
```

**5. Create a histogram of the patient ages in months**

```{r}
patient_age_histogram <- citbi |>
    ggplot(aes(x = age_in_month)) +
    geom_histogram(binwidth = 8, color = "black", fill = "pink") +
    labs(
        title = "The Histogram of Patient Ages",
        subtitle = "PECARN Emergency Departments",
        x = "Ages (In Months)",
        y = "Count") +
    theme_minimal()

patient_age_histogram
```

*Interesting patterns noticed:* 
1. The patients’ age distribution is **right-skewed**, with a high concentration of infants, particularly between **0–40 months**, and a secondary increase during adolescence (approx. **150–200 months**, or **12–17 years**).
2. This secondary rise may be related to:
a. sports-related injuries,
b. increased risk-taking behavior, and
c. greater independence from parental supervision.
3. There's **a gradual decline in patient counts after early infancy (around 16 months) through approx. 150 months**, followed by **a sharp drop after 200 months** (likely because of the study age cutoff).

**6. Create a grouped summary table that shows the total count of patients for every combination of the loss of consciousness length and ciTBI outcome columns**

```{r}
loclen_citbi <- citbi |>
    group_by(loss_of_consciousness_len, citbi_outcome) |> # 13 groups
    summarize(count = n()) |> # count
    ungroup()

loclen_citbi
```

*Notice:*
- Longer duration of loss of consciousness (LOC) is associated with a higher risk of traumatic brain injury.
Specifically, the number of patients with a positive outcome increases with LOC duration:
< 5 seconds: 8 TRUE cases
5 seconds to < 1 minute: 25 TRUE cases
1–5 minutes: 30 TRUE cases
> 5 minutes: 90 TRUE cases (highest)

**7. Create a bar chart to visualize for each category side-by-side in length of loss of consciousness by ciTBI outcome**
*Idea:*
- x-axis: LOC
- y-axis: counts of patients
- geom_bar(): automatically counts (no y needed)
- Bars grouped by ciTBI outcome (TRUE vs FALSE) #fill does this job!

```{r}
LOC_by_ciTBI_bar <- citbi |>
    drop_na(loss_of_consciousness_len, citbi_outcome) |> # drop NA values
    mutate(citbi_outcome = ifelse(citbi_outcome, "ciTBI", "No ciTBI")) |>
    ggplot(aes(x = loss_of_consciousness_len, fill = citbi_outcome)) + 
    geom_bar(position = "dodge", color = "black") + #position = "dodge": no stack bars! (but side-by-side)
    labs(
        title = "Length of Loss of Consciousness by ciTBI outcome",
        subtitle = "PECARN Emergency Departments",
        x = "Length of Loss of Consciousness (LOC)",
        y = "Count",
        fill = "ciTBI Outcome"
    ) +
    theme_minimal()
LOC_by_ciTBI_bar
```

**8. Create 3 visualizations for the relationship between total GCS score, Age in years, and ciTBI outcome**
- GCSTotal ~ Glasgow Coma Scale score, a neurological assessment used to evaluate the patient's level of consciousness.

```{r}
# Age in Years:
citbi <- citbi |>
    mutate(age_in_year = as.integer(age_in_month / 12))

citbi
```

*a. A stacked normalized bar chart to visualize how the proportion of patients with ciTBI varies across different ages*

```{r}
ciTBI_by_age <- citbi |>
    drop_na(citbi_outcome) |>
    mutate(citbi_outcome = ifelse(citbi_outcome, "ciTBI", "No ciTBI")) |>
    ggplot(aes(x = age_in_year, fill = citbi_outcome)) +
    geom_bar(position = "fill", color = "black") + #position = "fill": normalizes each bar to 100% -> each bar shows proportion, not raw count (default)
    labs(
        title = "Proportion of patients with ciTBi by Age",
        subtitle = "PECARN Emergency Departments",
        x = "Age (in Years)",
        y = "Proportion",
        fill = "ciTBI Outcome"
    ) +
    theme_minimal()

ciTBI_by_age
```

*b. A stacked normalized bar chart shows the proportion of patients with ciTBI across age, now faceted by total GCS score*
- To explore whether the relationship between age and ciTBI differs across Glasgow Coma Scale scores.

```{r}
ciTBI_by_age_facet <- citbi |>
    drop_na(citbi_outcome, GCSTotal) |>
    mutate(citbi_outcome = ifelse(citbi_outcome, "ciTBI", "No ciTBI")) |>
    ggplot(aes(x = age_in_year, fill = citbi_outcome)) +
    geom_bar(position = "fill", color = "black") +
    facet_wrap(~ GCSTotal) +
    labs(
        title = "Proportion of patients with ciTBi by Age",
        subtitle = "PECARN Emergency Departments",
        x = "Age (in Years)",
        y = "Proportion",
        fill = "ciTBI Outcome"
    ) +
    theme_minimal()

ciTBI_by_age_facet 
```

*c. A stacked bar chart shows the number of patients with ciTBI across age faceted by total GCS score*

```{r}
ciTBI_by_age_count <- citbi |>
    mutate(citbi_outcome = ifelse(citbi_outcome, "ciTBI", "No ciTBI")) |>
    ggplot(aes(x = age_in_year, fill = citbi_outcome)) +
    geom_bar(color = "black") +
    facet_wrap(~ GCSTotal, scales = "free_y") + #scales = "free_y": it lets each facet scale independently (for easier exploration)
    labs(
        title = "Number of patients with ciTBi by Age",
        subtitle = "PECARN Emergency Departments",
        x = "Age (in Years)",
        y = "Count",
        fill = "ciTBI Outcome"
    ) +
    theme_bw()

ciTBI_by_age_count
```

**9. Create a scatter plot to visualize any two numeric variables in the dataset**
- I chose to visualize GCSTotal with age_in_year, colored by citbi_outcome.
```{r}
GCS_score_by_Age_ciTBI <- citbi |>
    drop_na(age_in_year, GCSTotal, citbi_outcome) |>
    mutate (citbi_outcome = ifelse(citbi_outcome, "ciTBI", "No ciTBI")) |>
    ggplot(aes(
        x = age_in_year, 
        y = GCSTotal, 
        color = citbi_outcome)) +
    geom_jitter(width = 0.2, height = 0.2, alpha = 0.4) + #or geom_point(position = "jitter", alpha = 0.4)
    labs(
        title = "GCS Score by Age",
        subtitle = "Colored by ciTBI Outcome",
        x = "Age (in Years)",
        y = "Total GCS",
        color = "ciTBI Outcome"
    ) +
    theme_minimal()
```

*Interesting patterns noticed:* 
1. Lower GCS scores are associated with a higher likelihood of ciTBI.
2. Higher GCS scores are strongly associated with no ciTBI.
3. There is strong clustering of patients at high GCS scores, indicating most patients present with mild neurologic impairment.
4. GCS scores <8 show a noticeably higher concentration of ciTBI cases.
5. GCS scores >12 are dominated by non-ciTBI cases, with ciTBI events being rare.
6. Age shows no clear separation between ciTBI and non-ciTBI cases, suggesting that GCS score is a stronger indicator of ciTBI risk than age & age alone does not have much valuable predictive value.

**10. Create a table that summarizes at least 2 statistics grouped by 2 categorical variables**
- I chose 2 categorical variables: gender and death_tbi.
- **My question:** "Is there a difference in TBI-related death rates by gender?"

```{r}
death_by_gender <- citbi |>
    drop_na(gender, death_tbi) |>
    group_by(gender, death_tbi) |> #grouped by 2 categorical variables
    summarise(
        count = n(), #count number of patients: 4 rows x 3 columns
        .groups = "drop" #ungroup
    ) |>
    group_by(gender) |>
    mutate(
        proportion = count / sum(count) #death rate within gender
    )
```

*Interesting patterns noticed:* 
1. TBI-related death is rare in both genders, with mortality rates well below 1%; specifically, the proportion is approximately 0.0013 in males and 0.0023 in females.
2. Females show a slightly higher proportion of TBI-related deaths compared to males.
3. Overall case counts differ by gender, with more male patients than female patients, while the absolute number of deaths is similar between the 2 groups.

## Extra:
- I realized that I was interested in identifying patterns related to **gender** and wanted to explore its relationship with **other categorical variables**. 
- To do this more efficiently, I built a reusable function that allows me to examine patterns between gender and any categorical variable simply by supplying the variable name.

```{r}
gender_related <- function(categorical_variable){
    citbi |>
    drop_na(gender, {{ categorical_variable }} ) |>
    group_by(gender, {{ categorical_variable }}) |>
    summarise(
        count = n(),
        .groups = "drop"
    ) |>
    group_by(gender) |>
    mutate(
        proportion = count / sum(count)
    )
}
```

```{r}
#Check to see if my function right:
gender_related(neurological_deficit) #Now, it works as expected!
```


## Save all the data products for report.qmd:
```{r}
saveRDS(patient_age_histogram, "patient_age_histogram.rds")
saveRDS(LOC_by_ciTBI_bar, "LOC_by_ciTBI_bar.rds")
saveRDS(citbi, "citbi.rds")
saveRDS(ciTBI_by_age_facet, "ciTBI_by_age_facet.rds")
saveRDS(GCS_score_by_Age_ciTBI, "GCS_score_by_Age_ciTBI.rds")
```