---
title: "Pediatric ciTBI ER Visits: Key EDA Findings"
author: "Kartlyn Nguyen"
format: typst
---

```{r}
#| include: false
library(tidyverse)
citbi <- readRDS("RDS_files/citbi.rds")
```

## Data Products & Key Findings

### 1. Age Distribution of Pediatric Head Trauma Patients

-   **Questions answered:** "Who is in the dataset? What age ranges dominate pediatric head trauma cases in PECARN?"

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 3
patient_age_histogram <- readRDS("RDS_files/patient_age_histogram.rds")

patient_age_histogram
```

~[**Description:**]{.underline}*This plot shows which age groups most commonly present with head trauma, highlighting high-risk periods for targeted prevention and care.*~

#### Key Takeaways:

-   **Right-skewed**, with the highest concentration among infants (0 - approx. 50 months). A slight, secondary increase occurs during adolescence (150-200 months).
-   A gradual decline in patient counts after early infancy (around 16 months) through approx. 150 months & The sharp drop at older ages likely reflects the study age cutoff.

### 2. Proportion of Patients with ciTBI by Age (Faceted by Total GCS Score)

-   **Questions answered:** "For patients with similar GCS scores, does age meaningfully change ciTBI risk?"

```{r}
#| echo: false
#| fig-width: 6
#| fig-height: 3
ciTBI_by_age_facet <- readRDS("RDS_files/ciTBI_by_age_facet.rds")

ciTBI_by_age_facet
```

~**Description:***This\ faceted\ plot\ shows\ the\ proportion\ of\ patients\ with\ and\ without\ ciTBI\ across\ ages,\ with\ each\ panel\ representing\ a\ different\ total\ Glasgow\ Coma\ Scale\ (GCS)\ score.*~

#### Key Takeaways:

-  When neurological severity (GCS) is held constant, **age does not substantially change the proportion of ciTBI**, while **lower GCS scores consistently show higher ciTBI risk**.

### 3. GCS Score by Age, Colored by ciTBI Outcome

-   **Questions answered:** "How do GCS scores differ between patients with and without ciTBI?"

```{r}
#| echo: false
#| fig-width: 6
#| fig-height: 3
GCS_score_by_Age_ciTBI <- readRDS("RDS_files/GCS_score_by_Age_ciTBI.rds")

GCS_score_by_Age_ciTBI 
```

~[**Description:**]{.underline}*This scatter plot displays the relationship between patient age and GCS Total Score, with dots colored by ciTBI outcome.*~

#### Key Takeaways:

- **Higher GCS scores are dominated by non-ciTBI cases**.
-   **Age does not clearly distinguish ciTBI from non-ciTBI cases**, suggesting **GCS score is a stronger predictor of ciTBI risk** than age alone.

### 4. Loss of Consciousness (LOC) Length and ciTBI Risk

-   **Questions answered:** "Do patients with clinically important traumatic brain injury experience longer loss of consciousness than those without ciTBI?"

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 3
LOC_by_ciTBI_bar <- readRDS("RDS_files/LOC_by_ciTBI_bar.rds")

LOC_by_ciTBI_bar
```

~[**Description:**]{.underline}*This plot compares the distribution of loss of consciousness (LOC) duration between patients with and without ciTBI.*~

#### Key Takeaways:

-   ciTBI risk increases with longer LOC duration.
-   Patients with **LOC \>5 minutes** show the highest number of ciTBI cases (with 90 TRUE cases).

### 5. Proportion of ciTBI by Clavicle Trauma Status

-   **Questions answered:** “Is ciTBI more likely when clavicle trauma is present?”

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 3
clavicle_ciTBI <- citbi |>
    drop_na(clavicle_trauma, citbi_outcome) |>
    group_by(clavicle_trauma) |>
    summarize(
        count = n(),
        ciTBI_rate = mean(citbi_outcome == 1)
    ) |>
    ggplot(aes(x = clavicle_trauma, y = ciTBI_rate)) +
    geom_col(fill = "#f5a753ff", width = 0.5) +
    scale_y_continuous(labels = scales::percent_format()) +
    labs(
        title = "Proportion of ciTBI by Clavicle Trauma Status",
        subtitle = "PECARN Emergency Departments",
        x = "Clavicle Trauma",
        y = "Proportion with ciTBI"
        ) +
    theme_minimal(base_size = 12)

clavicle_ciTBI 
```

~***Description:** This figure shows the relationship between the presence of clavicle trauma and ciTBI outcome.*~

#### Key Takeaways:

-   The proportion of ciTBI is higher among patients with clavicle trauma, suggesting an association with increased injury severity.